Corpus Experiments for Word Embeddings

### 2015-08-25 article/main.tex
fix heatmap whitespace; rework WFVE results section; tables

\textcolor{red}{what about some work introducing tokens into a corpus for experiments?}

@@

\subsection{Results}
\textcolor{red}{rework}
For well-learned representations, word frequency has no effect on word vector direction.
Figure \ref{word-frequency-experiment-heatmap} shows the cosine similarity of the pairs of word vectors for the tokens introduced into the corpus.
Recall that the cosine similarity measures the extent to which two vectors have the same direction, taking a maximum value of $1$ and a minimum value of $-1$.
Notice that the word vectors associated to tokens derived from a word other that \word{the} and \word{VOID} have the same direction.
The same is true of the word vectors associated to the \word{VOID\_i} for $i$ \textcolor{red}{below $7$ or $8$}, but that the word vectors \word{VOID\_i} gradually diverge directionally for higher values of $i$.
This demonstrates the difficulty of learning the co-occurrence distribution of $\word{VOID}$, the global frequency distribution.
Because of its noisiness, a higher number of samples (occurrences) is required to learn this co-occurrence distribution.
The same holds, but to a less pronounced extent, for the co-occurrence distribution of the stopword \word{the}.
\textcolor{red}{The poorly learned representations for \word{the\_i} for $i = 15, \dots $ and for \word{VOID\_i} for $i = 7, \dots$ are therefore disregarded.}

\textcolor{red}{include a table with nearest neighbours in the normal vocab for the to support this argument}
Table \ref{word-frequency-experiment-cosine-similarity}

It remains, therefore, to consider the effect of frequency variation on word vector length.
Figure \ref{fig:word-frequency-experiment-graph}, on the other hand, shows this relationship for individual words, both for the word vectors, ``syn0'' and for the discarded second layer vectors ``syn1neg'' (included for completeness).
Each line corresponds to a single word, and the points on each line indicate the frequency and vector length for the tokens derived from the corresponding word.
For example, the six points on the line corresponding to \word{protestant} are labeled, from right to left, by the tokens \word{PROTESTANT\_1}, \word{PROTESTANT\_2}, \dots, \word{PROTESTANT\_6}.
The number of points on the line is determined by the frequency of the original word.
For example, the frequency of the \word{protestant} can be halved at most $6$ times and remain about the minimum frequency cut-off.

Because all the points on a line share a co-occurrence distribution (only the frequency varies), figure \ref{fig:word-frequency-experiment-graph} demonstrates conclusively that length does indeed depend on frequency directly, and not only via some correlation between word frequency and word significance.
Moreover, this relationship is linear for any particular word.
Notice also that the relative positions of the norms of the word vectors associated to the test words is independent of the frequency band.
Therefore these lines trace lines along which a frequency-adjusted norm would be constant.

Notice also that the word vector for \word{VOID} is non-zero.

@@

\begin{subsection}{Results}
Figure \ref{fig:co-occurrence-noise-heatmap} shows the cosine similarity of the pairs of word vectors for the tokens introduced into the corpus in this experiment.
Notice that the word vectors associated to tokens derived from the same word all share a common direction.
We conclude that co-occurrence noise has no effect on word vector direction.

@@

The co-occurrence distribution of \word{VOID} is the global frequency distribution, and in this sense pure background noise.
Thus the word vector of \word{VOID} is a special point in the feature space.
Figure \ref{fig:frequency-norm-graph} shows that this point is not at the origin of the feature space (i.e. is not the zero vector).
The origin, however, is implicitly the point of reference in the word2vec word similarity task.
This raises the question of whether improved performance on the similarity task could be achieved by transforming the feature space (or modifying the model) such that the representation of pure noise (i.e. the vector for \word{VOID}) was at the origin of the transformed feature space. 
\end{subsection}

>>>>

\textcolor{red}{what about some work introducing tokens into a corpus for experiments?}
\textcolor{red}{cite ``contextual distinctiveness: a new lexical property ...'' by McDonald and Shillcock?}

@@

\subsection{Results: word frequency and vector direction}\label{WFVE-direction}
Figure \ref{word-frequency-experiment-heatmap} shows the cosine similarity of the pairs of word vectors for the tokens introduced into the corpus.
Recall that the cosine similarity measures the extent to which two vectors have the same direction, taking a maximum value of $1$ and a minimum value of $-1$.
The number of word vectors associated to any word is the number of times that its frequency cat be halved and remain above the chosen minimum word count threshold of $128$.

Consider first the words other than \word{the} and \word{VOID}, e.g. \word{protestant}.
These words occur in a limited number of contexts, and so fewer samples are required to capture their co-occurrence distributions of such words.
Thus their word vectors can be learned with less samples.
It is therefore to be expected that the word vectors of their associated tokens can be properly learned from the modified corpus.
The red blocks of figure \ref{word-frequency-experiment-heatmap} clearly show that the direction of these word vectors depends only upon the original word.

The word \word{VOID}, on the other hand, occurs with probability in every context, by construction.
Thus a much higher number of samples are required to capture its co-occurrence distribution, and thereby to learn its word vector.
The same is true, but to a lesser extent, of the stopword \word{the}.
The ``difficulty'' of learning the word vector associated to \word{VOID\_i} or \word{the\_i} increases with $i$.
This is consistent with figure \ref{word-frequency-experiment-heatmap}, where it is apparent that the word vectors associated to \word{VOID\_i} for $i = 1, \dots, 5$ all share a common direction, while the word vectors for diverge directionallly for higher values of $i$.
The same trend is apparent, but less extremely so, for \word{the}.
The word vectors for \word{the\_i} share a common direction for $i = 1, \dots, 14$, before diverging directionally for higher values of $i$.
We conclude that these directional divergences are artifacts of poor learning due to an insufficient number of occurrences.
Table \ref{word-frequency-experiment-cosine-similarity}, showing the most similar words to the word vector of each \word{the\_i} provides further support for this claim.
In further discussion, the tokens \word{VOID\_i} (respectively, \word{the\_i}) whose word vectors have a cosine similarity less than $0.8$ with \word{VOID\_1} (respectively, \word{the\_1}) are therefore disregarded.

In conclusion, word frequency has no effect on word vector direction, given a sufficient number of samples to learn the co-occurrence distribution of the word.
%For well-learned representations, word frequency has no effect on word vector direction.

\subsection{Results: word frequency and vector length}
It remains, therefore, to consider the effect of frequency variation on word vector length.
Figure \ref{fig:word-frequency-experiment-graph} shows this relationship for individual words, both for the word vectors, ``syn0'' and for the discarded second-layer vectors ``syn1neg'' (included for completeness).
Each line corresponds to a single word, and the points on each line indicate the frequency and vector length for the tokens derived from the corresponding word.
For example, the six points on the line corresponding to \word{protestant} are labeled, from right to left, by the tokens \word{PROTESTANT\_1}, \word{PROTESTANT\_2}, \dots, \word{PROTESTANT\_6}.
Again, the number of points on the line is determined by the frequency of the original word.
For example, the frequency of the \word{protestant} can be halved at most $6$ times and remain about the minimum frequency cut-off.

Because all the points on a line share a co-occurrence distribution (only the frequency varies), figure \ref{fig:word-frequency-experiment-graph} demonstrates conclusively that length does indeed depend on frequency directly, and not only via some correlation between word frequency and word significance.
Moreover, this relationship is linear for any particular word.
Notice also that the relative positions of the norms of the word vectors associated to the experiment words is independent of the frequency band, i.e. the plotted lines rarely cross.
The plotted lines may thus be seen as the approximate contours of a hypothetical measure of word vector length normalised for word frequency effects.

Notice the inflection of the ``syn1neg'' curve for \word{the} at approximately $10^6$.  We have no explanation of this.

As an aside, notice also that the word vector length for the tokens \word{VOID\_i} is approximately constant.
Since the direction is also constant (cf. subsection \ref{WFVE-direction}), it is therefore sensible to speak of the word vector of \word{VOID} irrespective of its frequency.
In particular, \word{VOID\_1} may be taken as an approximation.

@@

\begin{subsection}{Results}
	%The co-occurrence distributions of e.g. the tokens \word{}, for $i=1, \dots, 8$ increasingly approximate the global word frequency distribution, i.e. co-occurrence distribution of the token \word{VOID}.
	%Since the vectorisation of \word{VOID} is independent of its frequency (cf. remark \ref{}), the word vectors of \word{} can therefore be considered as a series of intermediate points on a hypothetical path in the feature space beginning at the word vector for \word{} and ending at the word vector for \word{VOID}.
	\textcolor{red}{all of the following is true (using the heatmap that i say it does) but it ignores the artifact that the path converges to the word vector for \word{VOID}, not to zero -- this is fine to the extent that they are approximately the same.}

Figure \ref{fig:co-occurrence-noise-heatmap} shows the cosine similarity of the pairs of word vectors for the tokens introduced into the corpus in this experiment.
Notice that the word vectors associated to tokens derived from the same word all share a common direction.
We conclude that co-occurrence noise has no effect on word vector direction.

@@

The co-occurrence distribution of \word{VOID} is the global frequency distribution, and in this sense pure background noise.
Thus the word vector of \word{VOID} is a special point in the feature space.
Figure \ref{fig:frequency-norm-graph} shows that this point is not at the origin of the feature space (i.e. is not the zero vector).
The origin, however, is implicitly the point of reference in the word2vec word similarity task.
This raises the question of whether improved performance on the similarity task could be achieved by transforming the feature space (or modifying the model) such that the representation of pure noise (i.e. the vector for \word{VOID}) is at the origin of the transformed feature space.

\textcolor{red}{can it be argued that the feature space must be fundamentally euclidean, given that the path marked by the vectorisations of intermediate words is a straight line? i.e. if such paths maybe considered as ``geodesics'' between words, and we show that these geodesics appear as straight lines when an euclidean metric is imposed, then suggests that the euclidean metric is in fact a good choice.  what would happen if we chose another metric?}
